---
title: "Differential Expression for Control vs Heat Shock HeLa cells"
output: html_notebook
---

### Some notation notes:
* cnt -> control (37ºC)
* OR -> heat shock (42ºC) invitro

*heat shock was consistent among tissues*

### DNA to protein:
ATG -> AUG -> M

*multiple codons can make the same amino acid*

*what overexpress in response to heat shock*

* VST normalize the counts (why VST?)
* need to account for tissues that have differing amounts of high DEGs
* bar graph for DEG big picture among tissues
* ppi
* this is causal analysis
* volcanao plot (higher genes on y axis are more DE)
* why do we only care about the overexpressed genes?
* promoter and protein bind and make transcription happen?
* ODE model of cellular stress response

## install required packages
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install(c("DESeq2", "topGO", "fgsea", "clusterProfiler", "org.Hs.eg.db"))
install.packages(c("ggplot2", "pheatmap", "RColorBrewer", "EnhancedVolcano", "tidyverse"))
```

## libraries
```{r}
library("DESeq2")
library("tidyverse")
library("dplyr")
```


## read the csv to get count dataframe
```{r}
filename <- "/Users/emmagomez/code/EAGER/rna-seq-r/counts/read_counts.csv"
counts <- read.csv(filename)
```

## show columns
```{r}
colnames(counts)
```

## drop irrelevant columns
```{r}
rownames(counts) <- counts$gene_id
counts <- dplyr::select(counts, "HeLa1cnt", "HeLa2cnt", "HeLa3cnt",
                        "HeLa1_0R", "HeLa2_0R", "HeLa3_0R")
head(counts)
```


## change counts to integers
```{r}
to_mutate <- colnames(counts)[1:length(colnames(counts))]
counts <- counts %>%
  mutate(across(all_of(to_mutate), as.integer))
```

## metadata
```{r}
condition <- factor(c(rep("control", 3), rep("heat_shock", 3)))
col_data <- data.frame(
  row.names = colnames(counts),
  condition = condition
)
```


## construct DESeq2 object
```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = col_data,
                              design = ~condition)
```


## view results
```{r}
dds <- DESeq(dds)
res <- results(dds)
```

